---
title: "Weekl 7 Daily Challenge"
author: "Kenneth I. D."
date: "2/4/2020"
output:
  word_document: default
  html_document: default
---

**Question 1: What is the difference between the art and science of data analysis?**

**My Answer**
The _art_ of data science is the skill involved in knowing what insight to draw through critical thinking from a data set while the _science_ of data analysis involves knowing what methods and procedures to employ in applying statistics in the real world.

**Question 2: Suppose that the data for analysis includes the attribute age. The age values for the data tuples are (in increasing order) 13, 15, 16, 16, 19, 20, 20, 21, 22, ,22, 25, 25, 25, 25, 30, 33, 33, 35, 35, 35, 35, 36, 40, 45, 46, 52, 70.**

**(a.1) what is the mean of the data? (a.2) What is the median?**

```{r, echo=FALSE}
age <- c(13, 15, 16, 16, 19, 20, 20, 21, 22, 22, 25, 25, 25, 25, 30, 33, 33, 35, 35, 35, 35, 36, 40, 45, 46, 52, 70)

tab_age=table(age)
tab_age
df_age=as.data.frame(tab_age)
names(df_age)[1]='mode'
df_age

mean(age)
median(age)

paste("The Mean is", mean(age))
paste("The Median is", median(age))
```

**(b) What is the mode of the data? Comment on the modality (ie bimodal, trimodal etc)**

**Answer**
From the table of the data frame given above, 
```{r, echo=FALSE}
tab_age=table(age)
tab_age
df_age=as.data.frame(tab_age)
names(df_age)[1]='mode'
df_age
paste("The data is BIMODAL and its mode are 25 and 35")

```

**(c) What is the midrange of the data?**

```{r, echo=FALSE}
mean(range(age))
```

**(d) Can you find (roughly) the first quartile(Q1) and the third quartile(Q3) of the data?**
```{r, echo=FALSE}
Q1 <- quantile(age, 0.25)
Q3 <- quantile(age, 0.75)
paste("Q1 is", Q1)
paste("Q3 is", Q3)
```
**(e) Give the five number summary.**
```{r, echo=FALSE}
summary(age)
```
**(f) Show a boxplot of the data**

```{r, echo=FALSE}
boxplot(age)
```
**(g) How is a quantile-quantile plot different from a quantile plot?**
 A quantile plot is a graphical method used to show the approximate percentage values below or equal to the independent variable in a univariate distribution. These values are then plotted against their quantile. 
 While a quantile-quantile plot plots the quantiles of one univariate distribution agains the quantiles of another univariate distribution.

**Question 3: Suppose that a hospital tested the age and body fat for 18 randomly selected adults with the following results:**

```{r, echo=FALSE}
Age <- c(23, 23, 27, 27, 39, 41, 47, 49, 50, 52, 54, 54, 56, 57, 58, 58, 60, 61)
fat <- c(9.5, 26.5, 7.8, 17.8, 31.4, 25.9, 27.4, 27.2, 31.2, 34.6, 42.5, 28.8, 33.4, 30.2, 34.1, 32.9, 41.2, 35.7)


df_test=data.frame(Age, fat)
df_test
```
**(a) Calculate the mean, median and standard deviation of age and %fat**

Performing the five number summary to show the mean and the median
```{r, echo=FALSE}
summary(df_test)
sd_age <- sd(Age)
sd_fat <- sd(fat)
paste("The standard deviation of the age is", sd_age)
paste("The standard deviation of the percentage body fat is", sd_fat)
```

**(b)Draw the boxplots for age and %fat.**

```{r, echo=FALSE}
boxplot(Age, xlab='Age', ylab='Count')
```

```{r, echo=FALSE}
boxplot(fat, xlab='Fat', ylab='count')
```

**(c)Draw the scatter plot**
```{r, echo=FALSE}
plot(Age)
```


```{r, echo=FALSE}
plot(fat)
```

```{r, echo=FALSE}
plot(Age, fat)
```